---
title: "Testing things out for the final project"
author: "Youssra Yemmas"
date: "2023-12-08"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, message = FALSE}
library(tidyverse)
library(janitor)
library(knitr)
library(tidyr)
library(readxl)
library(lubridate)
library(readr)
library(stringr)
library(ggplot2)
library(tsibble)
library(dbplyr)
library(stats)
library(plotly)
library(ggridges)
library(forecast)
library(patchwork)
library(dplyr)
library(flexdashboard)
library(leaflet)
library(shiny)

library(leaflet.providers)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 7,
  out.width = "90%"
)


theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_color_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```
## Describing the motivation for our project
We like many New Yorkers see rats as a major problem that has only worsened following the pandemic. New York City government agrees and has implemented and promoted a flashy new initiative they are calling "Send Rats Packing." https://www.nyc.gov/assets/queenscb2/downloads/pdf/notices/2023/SetoutTimes-Residential-Flyer-2023-02.pdf This initiative is mainly composed of a new rule involving trash that aims to reduce the time that trash, recycling, and curbside composting will sit on the sidewalk. The new rule went into effect on April 1, 2023 and left Residential buildings and their managers with two options -Place waste out after 6:00 PM in a container of 55 gallons or less with a secure lid or Place waste out after 8:00 PM, if putting bags directly on the curb. Although everyone wants to see rats gone not everyone is on board with this new rule and many question if it is actually going to result in less rats. The NYC Building Supers an organization composed of building maintenance workers like porters and supers across the 5 boroughs has called this rule "outrageous and unfair" which requires them to work 14 hour day just to comply with the new rule. They have banded together to strike this rule by engaging in city hall protests and acts of civil disobedience by not complying with the new trash time. https://nycbuildingsupers.com/ Given this backdrop and the general skepticism of the effectiveness of the measure we were motivated to explore whether or not this trash time is effective at reducing the presence of rats across the 5 boroughs. 
## Cleaning rat sightings dataset

```{r import and clean obs, message = FALSE, warning=FALSE}

rat_sightings = 
  read_csv ("./data/Rat_Sightings.csv") |>
  janitor::clean_names(case = "snake") |>
  separate(created_date, sep="/", into = c("month", "day", "year")) |> 
  separate(year, sep=" ", into = c("year")) |>
  filter(borough != "STATEN ISLAND") |> 
  filter(year %in% c("2019", "2020", "2021", "2022", "2023")) |>
  mutate(
    borough_id = recode(
      borough, 
      "MANHATTAN" = 1,
      "BRONX" =2,
      "BROOKLYN"=3,
      "QUEENS"= 4)) |>
  mutate(
    month = as.numeric(month),
    year = as.numeric(year)
  ) |>
  select(unique_key, month, day, year, location_type, incident_zip, borough, location, borough_id) |>
  mutate(
    borough = str_to_sentence(borough)
  )
```

### Data Cleaning for Waste Tonnage data
```{r}
waste_tonnage = read_csv("./data/DSNY_Monthly_Tonnage_Data_20231202.csv") %>%
  clean_names(case = "snake") %>%
  mutate(date_split = strsplit(month, "/")) %>%
  mutate(
    year = as.integer(sapply(date_split, function(x) x[1])),
    month = as.integer(sapply(date_split, function(x) x[2]))
  ) %>%
  filter(year %in% c(2022, 2023)) %>% 
  mutate(total_organics = resorganicstons + schoolorganictons)

waste_tonnage = waste_tonnage %>% 
  group_by(borough, month, year, borough_id) %>% 
  summarise(
    total_organics = sum(total_organics, na.rm = TRUE),
    total_refuse = sum(refusetonscollected, na.rm = TRUE)
    )
```

### Merging rat sightings and waste tonnage data
```{r}
rat_waste_merged = left_join(rat_sightings, waste_tonnage, by = c("borough_id", "month", "year", "borough"))
```



## filtering out the N/A's in a dataset
```{r}
rat_waste_filtered = subset(rat_waste_merged, !is.na(total_organics) & !is.na(total_refuse))
```

## plot 1: total trash by borough for 2022 and 2023
```{r}
waste_tonnage_by_borough = rat_waste_filtered |> 
  group_by(borough) |> 
  summarise(total_tonnage = sum(total_organics + total_refuse))

ggplot(waste_tonnage_by_borough, aes(x = reorder(borough, -total_tonnage), y = total_tonnage)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(x = "Borough", y = "Trash Tonnage", title = "Trash Tonnage by Borough, 2022 - 2023") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## bar graph by total refuse and total organics (had to separate b/c they are on a different scale)
```{r}
waste_by_borough_refuse <- aggregate(total_refuse ~ borough, data=waste_tonnage, sum)

# barplot for total refuse
barplot(waste_by_borough_refuse$total_refuse,
        main="Total Refuse Production by Borough",
        xlab="Borough",
        ylab="Total Refuse (tons)",
        col="blue",
        names.arg=waste_by_borough_refuse$borough,
        las=2)

waste_by_borough_organics <- aggregate(total_organics ~ borough, data=waste_tonnage, sum)

# barplot for total organics
barplot(waste_by_borough_organics$total_organics,
        main="Total Organics Production by Borough",
        xlab="Borough",
        ylab="Total Organics (tons)",
        col="green",
        names.arg=waste_by_borough_organics$borough,
        las=2)

```




## plot 2: trash by borough & month
```{r}
total_tonnage_by_borough_year_month = rat_waste_filtered |> 
  group_by(borough, year, month) |> 
  summarise(total_tonnage = sum(total_organics + total_refuse))

total_tonnage_by_borough_year_month$year_month <- as.Date(paste(total_tonnage_by_borough_year_month$year, total_tonnage_by_borough_year_month$month, "01", sep = "-"))

ggplot(total_tonnage_by_borough_year_month, aes(x = year_month, y = total_tonnage, group = borough, color = borough)) +
  geom_line() +
  labs(x = "Year-Month", y = "Amount of Trash (in tons)", title = "Trash Tonnage by Month and Borough") +
  scale_x_date(date_labels = "%Y-%m", date_breaks = "1 month") +
  scale_y_continuous(labels = scales::number_format(scale = 1e-6)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

##  statistical analysis 1: statistical test (ANOVA) to assess the differences in waste tonnage among the boroughs.
```{r}



anova_result_organics = aov(total_organics ~ borough, data = waste_tonnage) |> broom::tidy()

anova_result_refuse = aov(total_refuse ~ borough, data = waste_tonnage) |> broom::tidy()

```

The one-way ANOVA test was performed to assess the differences in organic waste tonnage among the boroughs. The ANOVA test revealed statistically significant differences in organic waste tonnage among the boroughs (F(4, 112) = 10.5, p < 0.05). 

A one-way ANOVA test was conducted to examine the differences in refuse waste tonnage among the boroughs.The ANOVA test revealed statistically significant differences in refuse waste tonnage among the boroughs (F(4, 112) = 82.1, p < 0.001).


### Data Cleaning for Waste Tonnage data
```{r}
waste_tonnage = read_csv("./data/DSNY_Monthly_Tonnage_Data_20231202.csv") %>%
  clean_names(case = "snake") %>%
  mutate(date_split = strsplit(month, "/")) %>%
  mutate(
    year = as.integer(sapply(date_split, function(x) x[1])),
    month = as.integer(sapply(date_split, function(x) x[2]))
  ) %>%
  filter(year %in% c(2022, 2023)) %>% 
  mutate(total_organics = resorganicstons + schoolorganictons)

waste_tonnage = waste_tonnage %>% 
  group_by(borough, month, year, borough_id) %>% 
  summarise(
    total_organics = sum(total_organics, na.rm = TRUE),
    total_refuse = sum(refusetonscollected, na.rm = TRUE)
    )
```

### Merging rat sightings and waste tonnage data
```{r}
rat_waste_merged = left_join(rat_sightings, waste_tonnage, by = c("borough_id", "month", "year", "borough"))
```



### One question we are interest in is how do the rat sightings differ over time and across boroughs
```{r}
rat_sightings = 
  read_csv ("./data/Rat_Sightings.csv") |>
  janitor::clean_names(case = "snake") |>
  separate(created_date, sep="/", into = c("month", "day", "year")) |> 
  separate(year, sep=" ", into = c("year")) |>
  filter(borough != "STATEN ISLAND") |> 
  filter(year %in% c("2019", "2020", "2021", "2022", "2023")) |>
  mutate(
    borough_id = recode(
      borough, 
      "MANHATTAN" = 1,
      "BRONX" =2,
      "BROOKLYN"=3,
      "QUEENS"= 4)) |>
  mutate(
    month = as.numeric(month),
    year = as.numeric(year)
  ) |>
  select(unique_key, month, day, year, location_type, incident_zip, borough, location, borough_id) |>
  mutate(
    borough = str_to_sentence(borough)
  )

summary(rat_sightings)
variable_types <- sapply(rat_sightings, class)
print(variable_types)
# variables are not classified well for analysis so will need to convert numeric variables
numeric_vars_to_convert <- c("unique_key", "month", "year", "incident_zip", "borough_id")

rat_sightings <- rat_sightings %>% 
  mutate(across(all_of(numeric_vars_to_convert), as.factor))
        
variable_types <- sapply(rat_sightings, class)
print(variable_types)

# number of rat sightings by boro each year
rats_boro = rat_sightings %>% 
  janitor::clean_names() %>% 
  select(borough, year, unique_key) %>% 
  group_by(borough, year) %>% 
  count() %>% 
  summarize(avg_rat_sightings = mean(n)) %>% 
  ungroup %>% 
  spread(key = year, value = avg_rat_sightings) %>% 
  filter(borough != 'Unspecified')# I want to remove the unsepcified

knitr::kable(rats_boro)
  
```

We can see from the kable output that there was quite a substantial jump in rat sightings from 2020 to 2021 in all of the boroughs. This may be another COVID phenomena as restaurants shifted to more outdoor dining which deposited more food waste and other things that attract rats onto the streets during the pandemic and after the pandemic when indoor dining became less feasible. https://apnews.com/article/rats-new-york-9dc65afa66a3535cba01b1ea866973a1#:~:text=NEW%20YORK%20(AP)%20%E2%80%94%20They,so%20did%20the%20city's%20rats.

### Are the differences we can see in average rate sighting across time and boroughs statistically significant? 
```{r}
# I will test the statistical difference of average rat sighting across boroughs and across time.


rat_sightings_agg = rat_sightings |> 
  group_by(year, borough, month) |> 
  filter(borough != "Unspecified") %>% 
  summarise(count = n())

anova_result = aov(count ~ factor(year) * factor(borough), data = rat_sightings_agg) |> broom::tidy()

anova_result_no_interaction = aov(count ~ factor(year) + factor(borough), data = rat_sightings_agg) |> broom::tidy()


# Print the summary to get F-statistic and p-value
anova1_summary <- summary(anova_result)
knitr::kable(anova1_summary)

anova2_summary <- summary(anova_result_no_interaction)
knitr::kable(anova2_summary)
```
From the ANOVA tests above we can see from the p-value of both tests being <0.001 we can reject the null hypotheses and conclude that at least one of the average rat sightings by borough and by year are statistically different.

### Distributions
```{r}
rat_sightings_agg <- rat_sightings_agg %>%
  mutate(common_date = paste0(year, "-", month),
         common_date = lubridate::ym(common_date))

# Plot rat sightings over time
rats_yr_plot <- rat_sightings_agg %>%
  ggplot(aes(x = common_date, y = count, group = year, color = year)) +
  geom_point(alpha = .3) +
  geom_smooth(se = FALSE) +
  facet_wrap(~ borough, scales = "free_y", ncol = 2) +  # Facet wrap by borough
  labs(
    title = "Rat Sightings Over Time",
       x = "Date",
       y = "Rat Sightings Count") +
  theme(text = element_text(size = 15), 
        axis.text.x = element_text(angle = 60, hjust = 1, size = 10)) +
  scale_colour_discrete("Year") +
  scale_x_date(date_breaks = "3 month", labels = function(x) format(x, "%b")) +
  scale_color_viridis_d(end = .8)
print(rats_yr_plot)

ggplotly(rats_yr_plot)


rat_sightings_tsibble <- rat_sightings_agg %>%
  mutate(common_date = paste0(year, "-", month),
         common_date = lubridate::ym(common_date)) %>%
  as_tsibble(index = common_date, key = borough) %>%
  select(borough, common_date, count)  # Include common_date in the select statement

# Time series plot
rats_time_series_plot <- rat_sightings_tsibble %>%
  ggplot(aes(x = common_date, y = count, color = borough, group = borough)) +
  geom_line() +
  labs(title = "Rat Sightings Time Series",
       x = "Date",
       y = "Rat Sightings Count",
       color = "Borough") +
  theme_minimal()

# Print the plot
print(rats_time_series_plot)
ggplotly(rats_time_series_plot)

borough_data <- rat_sightings_tsibble %>%
  filter(borough == "Brooklyn")  

# Convert data to a univariate time series
ts_data <- ts(borough_data$count, frequency = 12)  
# Time series analysis using STL decomposition
ts_analysis <- stlf(ts_data, h = 12)  

# Plot time series decomposition
autoplot(ts_analysis) +
  labs(title = "Time Series Decomposition",
       y = "Rat Sightings Count",
       color = "Components") +
  theme_minimal()

# Obtain forecasts
forecast_result <- ts_analysis %>%
  forecast(h = 12)  


# Plot forecasts
autoplot(forecast_result) +
  labs(title = "Rat Sightings Forecast",
       x = "Date",
       y = "Rat Sightings Count") +
  theme_minimal()


rat_sightings_tsibble2 <- rat_sightings_agg %>%
  mutate(common_date = paste0(year, "-", month, "-01"),
         common_date = lubridate::ymd(common_date),
         rule_impact = ifelse(common_date >= "2023-04-01", "After", "Before")) %>%
  as_tsibble(index = common_date, key = c(borough, rule_impact)) %>%
  select(borough, common_date, count, rule_impact)

# Plot to analyze the impact of the rule
rats_rule_impact_plot <- rat_sightings_tsibble2 %>%
  ggplot(aes(x = common_date, y = count, color = rule_impact, group = interaction(borough, rule_impact))) +
  geom_line() +
  labs(title = "Impact of Rat Extermination Rule",
       x = "Date",
       y = "Rat Sightings Count",
       color = "Rule Impact",
       subtitle = "Comparison Before and After April 2013") +
  theme(text = element_text(size = 15), 
        axis.text.x = element_text(angle = 60, hjust = 1, size = 10)) +
  scale_colour_discrete("Year") +
  scale_x_date(date_breaks = "3 month", labels = function(x) format(x, "%b")) +
  scale_color_viridis_d(end = .8)

# Print the plot
print(rats_rule_impact_plot)
ggplotly(rats_rule_impact_plot)
```



### Visualizations
```{r}
viz1_data = rats_boro %>% 
  pivot_longer(cols = starts_with("20"),
               names_to = "Year",
               values_to = "avg_rat_sightings"
  )

  ggplot(viz1_data, aes(x = Year, y = avg_rat_sightings)) +
   geom_point(alpha = 0.3, size = 2) +
   geom_line(size = 1, alpha = 0.6) +
   facet_wrap(~borough, scales = "free_y") +
     theme(legend.position = "bottom",
         axis.text.y = element_text(color = "black", 
                                    size = 10,  hjust = 1), 
         axis.text.x = element_text(angle = 45, 
                                    hjust = 1, size = 10)) +
   labs(
     x = "Year",
     y = "Average Rat Sightings",
     title = "Average Rate Sightings From 2019-2023 by Borough"
   ) + 
     viridis::scale_colour_viridis() 
# I cannot seem to get the line to form so I will try it with a different data format 
  
 ggplot(rat_sightings_agg, aes(x = year, y = count)) +
   geom_point(alpha = 0.3, size = 2) +
   geom_line(size = 1, alpha = 0.6) +
   facet_wrap(~borough, scales = "free_y") +
     theme(legend.position = "bottom",
         axis.text.y = element_text(color = "black", 
                                    size = 10,  hjust = 1), 
         axis.text.x = element_text(angle = 45, 
                                    hjust = 1, size = 10)) +
   labs(
     x = "Year",
     y = "Average Rat Sightings",
     title = "Average Rate Sightings From 2019-2023 by Borough"
   ) + 
     viridis::scale_colour_viridis() 
 
 # This graph is also not what I had in mind 


```




