---
title: "Testing things out for the final project"
author: "Youssra Yemmas"
date: "2023-12-08"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, message = FALSE}
library(tidyverse)
library(tidyr)
library(readxl)
library(stringr)
library(dbplyr)
```
## Describing the motivation for our project
We like many New Yorkers see rats as a major problem that has only worsened following the pandemic. New York City government agrees and has implemented and promoted a flashy new initiative they are calling "Send Rats Packing." https://www.nyc.gov/assets/queenscb2/downloads/pdf/notices/2023/SetoutTimes-Residential-Flyer-2023-02.pdf This initiative is composed of a new rule involving trash that aims to reduce the time that trash, recyling, and curbside composting will sit on the sidewalk. The new went into effect on April 1, 2023 and left Residential buildings and their managers with two options -Place waste out after 6:00 PM in a container of 55 gallons or less with a secure lid or Place waste out after 8:00 PM, if putting bags directly on the curb. n
## Cleaning rat sightings dataset

```{r import and clean obs, message = FALSE, warning=FALSE}

rat_sightings = 
  read_csv ("./data/Rat_Sightings.csv") |>
  janitor::clean_names(case = "snake") |>
  separate(created_date, sep="/", into = c("month", "day", "year")) |> 
  separate(year, sep=" ", into = c("year")) |>
  filter(borough != "STATEN ISLAND") |> 
  filter(year %in% c("2019", "2020", "2021", "2022", "2023")) |>
  mutate(
    borough_id = recode(
      borough, 
      "MANHATTAN" = 1,
      "BRONX" =2,
      "BROOKLYN"=3,
      "QUEENS"= 4)) |>
  mutate(
    month = as.numeric(month),
    year = as.numeric(year)
  ) |>
  select(unique_key, month, day, year, location_type, incident_zip, borough, location, borough_id) |>
  mutate(
    borough = str_to_sentence(borough)
  )
```


```{r setup, include=FALSE}
library(tidyverse)
library(readr)
library(janitor)
library(dplyr)
```

### Data Cleaning for Waste Tonnage data
```{r}
waste_tonnage = read_csv("./data/DSNY_Monthly_Tonnage_Data_20231202.csv") %>%
  clean_names(case = "snake") %>%
  mutate(date_split = strsplit(month, "/")) %>%
  mutate(
    year = as.integer(sapply(date_split, function(x) x[1])),
    month = as.integer(sapply(date_split, function(x) x[2]))
  ) %>%
  filter(year %in% c(2022, 2023)) %>% 
  mutate(total_organics = resorganicstons + schoolorganictons)

waste_tonnage = waste_tonnage %>% 
  group_by(borough, month, year, borough_id) %>% 
  summarise(
    total_organics = sum(total_organics, na.rm = TRUE),
    total_refuse = sum(refusetonscollected, na.rm = TRUE)
    )
```

### Merging rat sightings and waste tonnage data
```{r}
rat_waste_merged = left_join(rat_sightings, waste_tonnage, by = c("borough_id", "month", "year", "borough"))
```

```{r setup, include=FALSE}
library(tidyverse)
library(readr)
library(janitor)
library(dplyr)
library(ggplot2)
library(stats)
library(stats)
```


## filtering out the N/A's in a dataset
```{r}
rat_waste_filtered = subset(rat_waste_merged, !is.na(total_organics) & !is.na(total_refuse))
```

## plot 1: total trash by borough for 2022 and 2023
```{r}
waste_tonnage_by_borough = rat_waste_filtered |> 
  group_by(borough) |> 
  summarise(total_tonnage = sum(total_organics + total_refuse))

ggplot(waste_tonnage_by_borough, aes(x = reorder(borough, -total_tonnage), y = total_tonnage)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(x = "Borough", y = "Trash Tonnage", title = "Trash Tonnage by Borough, 2022 - 2023") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## bar graph by total refuse and total organics (had to separate b/c they are on a different scale)
```{r}
waste_by_borough_refuse <- aggregate(total_refuse ~ borough, data=waste_tonnage, sum)

# barplot for total refuse
barplot(waste_by_borough_refuse$total_refuse,
        main="Total Refuse Production by Borough",
        xlab="Borough",
        ylab="Total Refuse (tons)",
        col="blue",
        names.arg=waste_by_borough_refuse$borough,
        las=2)

waste_by_borough_organics <- aggregate(total_organics ~ borough, data=waste_tonnage, sum)

# barplot for total organics
barplot(waste_by_borough_organics$total_organics,
        main="Total Organics Production by Borough",
        xlab="Borough",
        ylab="Total Organics (tons)",
        col="green",
        names.arg=waste_by_borough_organics$borough,
        las=2)

```




## plot 2: trash by borough & month
```{r}
total_tonnage_by_borough_year_month = rat_waste_filtered |> 
  group_by(borough, year, month) |> 
  summarise(total_tonnage = sum(total_organics + total_refuse))

total_tonnage_by_borough_year_month$year_month <- as.Date(paste(total_tonnage_by_borough_year_month$year, total_tonnage_by_borough_year_month$month, "01", sep = "-"))

ggplot(total_tonnage_by_borough_year_month, aes(x = year_month, y = total_tonnage, group = borough, color = borough)) +
  geom_line() +
  labs(x = "Year-Month", y = "Amount of Trash (in tons)", title = "Trash Tonnage by Month and Borough") +
  scale_x_date(date_labels = "%Y-%m", date_breaks = "1 month") +
  scale_y_continuous(labels = scales::number_format(scale = 1e-6)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

##  statistical analysis 1: statistical test (ANOVA) to assess the differences in waste tonnage among the boroughs.
```{r}



anova_result_organics = aov(total_organics ~ borough, data = waste_tonnage) |> broom::tidy()

anova_result_refuse = aov(total_refuse ~ borough, data = waste_tonnage) |> broom::tidy()

```

The one-way ANOVA test was performed to assess the differences in organic waste tonnage among the boroughs. The ANOVA test revealed statistically significant differences in organic waste tonnage among the boroughs (F(4, 112) = 10.5, p < 0.05). 

A one-way ANOVA test was conducted to examine the differences in refuse waste tonnage among the boroughs.The ANOVA test revealed statistically significant differences in refuse waste tonnage among the boroughs (F(4, 112) = 82.1, p < 0.001).


### Data Cleaning for Waste Tonnage data
```{r}
waste_tonnage = read_csv("./data/DSNY_Monthly_Tonnage_Data_20231202.csv") %>%
  clean_names(case = "snake") %>%
  mutate(date_split = strsplit(month, "/")) %>%
  mutate(
    year = as.integer(sapply(date_split, function(x) x[1])),
    month = as.integer(sapply(date_split, function(x) x[2]))
  ) %>%
  filter(year %in% c(2022, 2023)) %>% 
  mutate(total_organics = resorganicstons + schoolorganictons)

waste_tonnage = waste_tonnage %>% 
  group_by(borough, month, year, borough_id) %>% 
  summarise(
    total_organics = sum(total_organics, na.rm = TRUE),
    total_refuse = sum(refusetonscollected, na.rm = TRUE)
    )
```

### Merging rat sightings and waste tonnage data
```{r}
rat_waste_merged = left_join(rat_sightings, waste_tonnage, by = c("borough_id", "month", "year", "borough"))
```


```{r setup, include = FALSE}
library(tidyverse)
library(knitr)
library(janitor)
library(ggplot2)
library(plotly)


knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_color_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

### One question we are interest in is how do the rat sightings differ over time and across boroughs
```{r}
rat_sightings = 
  read_csv ("./data/Rat_Sightings.csv") |>
  janitor::clean_names(case = "snake") |>
  separate(created_date, sep="/", into = c("month", "day", "year")) |> 
  separate(year, sep=" ", into = c("year")) |>
  filter(borough != "STATEN ISLAND") |> 
  filter(year %in% c("2019", "2020", "2021", "2022", "2023")) |>
  mutate(
    borough_id = recode(
      borough, 
      "MANHATTAN" = 1,
      "BRONX" =2,
      "BROOKLYN"=3,
      "QUEENS"= 4)) |>
  mutate(
    month = as.numeric(month),
    year = as.numeric(year)
  ) |>
  select(unique_key, month, day, year, location_type, incident_zip, borough, location, borough_id) |>
  mutate(
    borough = str_to_sentence(borough)
  )

summary(rat_sightings)
variable_types <- sapply(rat_sightings, class)
print(variable_types)
# variables are not classified well for analysis so will need to convert numeric variables
numeric_vars_to_convert <- c("unique_key", "month", "year", "incident_zip", "borough_id")

rat_sightings <- rat_sightings %>% 
  mutate(across(all_of(numeric_vars_to_convert), as.factor))
        
variable_types <- sapply(rat_sightings, class)
print(variable_types)

# number of rat sightings by boro each year
rats_boro = rat_sightings %>% 
  janitor::clean_names() %>% 
  select(borough, year, unique_key) %>% 
  group_by(borough, year) %>% 
  count() %>% 
  summarize(avg_rat_sightings = mean(n)) %>% 
  ungroup %>% 
  spread(key = year, value = avg_rat_sightings) %>% 
  filter(borough != 'Unspecified')# I want to remove the unsepcified

knitr::kable(rats_boro)
  
```

We can see from the kable output that there was quite a substantial jump in rat sightings from 2020 to 2021 in all of the boroughs. This may be another COVID phenomena as restaurants shifted to more outdoor dining which deposited more food waste and other things that attract rats onto the streets during the pandemic and after the pandemic when indoor dining became less feasible. https://apnews.com/article/rats-new-york-9dc65afa66a3535cba01b1ea866973a1#:~:text=NEW%20YORK%20(AP)%20%E2%80%94%20They,so%20did%20the%20city's%20rats.

### Are the differences we can see in average rate sighting across time and boroughs statistically significant? 
```{r}
# I will test the statistical difference of average rat sighting across boroughs and across time.


rat_sightings_agg = rat_sightings |> 
  group_by(year, borough, month) |> 
  filter(borough != "Unspecified") %>% 
  summarise(count = n())

anova_result = aov(count ~ factor(year) * factor(borough), data = rat_sightings_agg) |> broom::tidy()

anova_result_no_interaction = aov(count ~ factor(year) + factor(borough), data = rat_sightings_agg) |> broom::tidy()


# Print the summary to get F-statistic and p-value
anova1_summary <- summary(anova_result)
knitr::kable(anova1_summary)

anova2_summary <- summary(anova_result_no_interaction)
knitr::kable(anova2_summary)
```



### Visualizations
```{r}
viz1_data = rats_boro %>% 
  pivot_longer(cols = starts_with("20"),
               names_to = "Year",
               values_to = "avg_rat_sightings"
  )

  ggplot(viz1_data, aes(x = Year, y = avg_rat_sightings)) +
   geom_point(alpha = 0.3, size = 2) +
   geom_line(size = 1, alpha = 0.6) +
   facet_wrap(~borough, scales = "free_y") +
     theme(legend.position = "bottom",
         axis.text.y = element_text(color = "black", 
                                    size = 10,  hjust = 1), 
         axis.text.x = element_text(angle = 45, 
                                    hjust = 1, size = 10)) +
   labs(
     x = "Year",
     y = "Average Rat Sightings",
     title = "Average Rate Sightings From 2019-2023 by Borough"
   ) + 
     viridis::scale_colour_viridis() 
# I cannot seem to get the line to form so I will try it with a different data format 
  
 ggplot(rat_sightings_agg, aes(x = year, y = count)) +
   geom_point(alpha = 0.3, size = 2) +
   geom_line(size = 1, alpha = 0.6) +
   facet_wrap(~borough, scales = "free_y") +
     theme(legend.position = "bottom",
         axis.text.y = element_text(color = "black", 
                                    size = 10,  hjust = 1), 
         axis.text.x = element_text(angle = 45, 
                                    hjust = 1, size = 10)) +
   labs(
     x = "Year",
     y = "Average Rat Sightings",
     title = "Average Rate Sightings From 2019-2023 by Borough"
   ) + 
     viridis::scale_colour_viridis() 
 
 # This graph is also not what I had in mind 


```


```{r setup, include=FALSE}
library(tidyverse)
library(ggridges)
library(patchwork)
library(dplyr)
library(flexdashboard)
library(leaflet)
library(shiny)

library(leaflet.providers)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 7,
  out.width = "90%"
)


```




```{r import and clean obs, message = FALSE, warning=FALSE}

rat_sightings = 
  read_csv ("./data/Rat_Sightings.csv") |>
  janitor::clean_names(case = "snake") |>
  separate(created_date, sep="/", into = c("month", "day", "year")) |> 
  separate(year, sep=" ", into = c("year")) |>
  filter(borough != "STATEN ISLAND") |> 
  filter(year %in% c("2019", "2020", "2021", "2022", "2023")) |>
  mutate(
    borough_id = recode(
      borough, 
      "MANHATTAN" = 1,
      "BRONX" =2,
      "BROOKLYN"=3,
      "QUEENS"= 4)) |>
  mutate(
    month = as.numeric(month),
    year = as.numeric(year)
  ) |>
  select(unique_key, month, day, year, location_type, incident_zip, borough, location, borough_id) |>
  mutate(
    borough = str_to_sentence(borough)
  )
```


```{r mapping dataset}
map_rat_sightings<-rat_sightings %>% 
  filter(year=="2022") %>% 
  separate(location, into = c("latitude", "longitude"), sep = ",", remove = FALSE) %>%  
mutate(
    longitude = str_replace(longitude, "\\)", ""),
    latitude = str_replace(latitude, "\\(", ""),
    longitude = as.numeric(longitude),
    latitude = as.numeric(latitude)) 
```



```{r}
ui <- bootstrapPage(
  titlePanel("NYC Rat Sightings by Month, 2022"),
  tags$style(type = "text/css", "html, body {width:100%;height:100%}"),
  leafletOutput("map", width = "100%", height = "600px"),
  absolutePanel(top = 10, right = 10,
                sliderInput("animation", "Month", min(map_rat_sightings$month), max(map_rat_sightings$month),
                            value = min(map_rat_sightings$month), step = 1,
                            animate = animationOptions(interval = 600, loop = FALSE)
                )
  )
)

server <- function(input, output, session) {
  
  filteredData <- reactive({
    subset <- map_rat_sightings[map_rat_sightings$month == input$animation, ]
    subset <- subset[complete.cases(subset$latitude, subset$longitude), ]
    subset
  })
  
  output$map <- renderLeaflet({
    pal <- colorFactor(palette = "Dark2", domain = filteredData()$borough) 
    
    leaflet(data = filteredData()) %>%
      addTiles() %>%
      addProviderTiles(providers$CartoDB.Positron) %>%
      fitBounds(
        min(filteredData()$longitude),
        min(filteredData()$latitude),
        max(filteredData()$longitude),
        max(filteredData()$latitude)
      ) %>%
      addCircleMarkers(
        lat = ~latitude,  
        lng = ~longitude, 
        radius = 3,              
        stroke = FALSE,
        fillOpacity = 0.8,
        color = ~pal(borough)    
      )
  })
}

shinyApp(ui, server)
```